Genetic Dilution erodes productivity: Exploring farmers' low adoption levels of improved maize in Ethiopia

### 2023-03-24 paper.tex
newest version

% 	\affil[1]{\small \emph{The World Bank}}
% 	\affil[2]{\small \emph{FAO}}
% 	\affil[3]{\small \emph{Assistant Research Professor at the Center for Data Science for Enterprise and Society, Cornell University}}
\author{%
 Aleksandr Michuda \footnote{Center for Data Science for Enterprise and Society, Cornell University}%
 \and Cristina Chiarella\footnote{Earth and Life Institute, UCLouvain, 1348 Louvain-la-Neuve, Belgium}%
 \and Oscar Barriga-Cabanillas\footnote{The World Bank}%
  \and Juan Sebastian Correa \footnote{Food and Agriculture Organization of the United Nations (FAO), 00153, Rome, Italy}%
  }

@@

    \clearpage\maketitle
	\thispagestyle{empty}
	\vspace*{-2em}
    \begin{center}\begin{abstract}
    \singlespacing
			\noindent  
			Despite large investments in developing new maize varieties, evidence shows that few farmers consistently adopt them. Moreover, average positive effects on yields mask considerable heterogeneity, particularly in contexts where seed are not traceable (missclasified as new varieties). We take a step further and study misclassification in the context of heterogeneous impacts to adoption. We use a three-year panel to estimate heterogeneous returns to adoption from self-reports of hybrid seed use and find null returns from adoption. We show how this result is driven misclassification of seeds that induce bias by exploiting a unique information on maize DNA-fingerprinting. When we account for seed misclassification, we find that heterogeneity in the actual quality of the genetic material of the seeds farmers categorized as improved varieties in the self-reported data masks positive returns to adoption, with the highest returns found for farmers using higher-purity germplasm, drought-tolerant maize, and newly released varieties. Our findings point to the wide dispersion of older and genetically diluted varieties, for which poorer farmers may be paying a premium, and therefor the need for policies that expand the accessibility of improved, context and geography-specific seeds, which display higher yields.

@@

Low adoption of hybrid seed, despite its benefits, has been a longstanding question in the study of agricultural technologies. Given their capacity to provide higher and less variable yields, lack of consistent adoption has received a lot of attention from the economic literature, with several explanations developed in the literature. These include lack of liquidity and other economic constraints \citep{Carter2014-fm}, growing conditions such as the type of soil, input use, and farmer characteristics \citep{Munshi2004-og}, imperfections in credit markets \citep{Croppenstedt2003-pq}, ambiguously defined property rights \citep{Place2000-el}, lack of commitment devices \citep{Duflo2009-iv}, high transportation costs that increase the cost of agricultural inputs \citep{Byerlee2013-qk}, and improved seed supply constraints \citep{Bird2020-nt}. An alternative mechanism for a lack of improvement has been the misclassification of seed and seed genetic dilution (\citealt{wossen2022misperceiving}; \citealt{kosmowski2019varietal}). Yield improvement and disease resistance wanes over time as the genetic material is recycled \citep{Abate2015-rj}, making that the net returns from these \textit{improved} seed is lower than expected, and decreases in the long term as farmers continue to use genetically dilute seed varieties, or ones that have been recycled over many generations (\citealt{euler2022because}; \citealt{wossen2019estimating}). Moreover, as genes from improved varieties enter the regular seed market its inadvertent use can potentially lead to inferior yields since they then to require the application of complementary inputs.

@@

The greater availability of DNA fingerprinting data makes possible a new revision of the role of misclassification of improved seed on the adoption of new varieties. For instance, \cite{wossen2019estimating} use DNA fingerprinting data to show that there is a 22\% bias in the productivity impact of improved cassava seed in Nigeria, suggesting that average impacts to adoption may not be as informative. Additionally, missclassification introduces a non-classical measurement error that can make that the observed productivity impact is attenuated or even reversed. This paper provides a way to account for misclassification of seed when additional, error-free data exists, such as DNA fingerprinting data. We estimate the comparative advantage of households directly using a Group Random Coefficients (GRC) Model and correct its heterogeneity estimates using misclassification rates from DNA fingerprinting data of the same plots \citep{Suri2011-oi, Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc}.

@@

Our results contribute to the literature on the drivers of low adoption rates of improved seed varieties as we are the first to consider how heterogeneous impacts to seed adoption are affected by misclassification of seed. The consensus in this literature is that the success of improved varieties depends on a range of factors, and that the widespread adoption of improved seeds is the result of complementary conditions and inputs that allow farmers to extract the full potential of these varieties, while other factos such as risk aversion \citep{Holden2016-vy}, farm size \citep{Ghimire2015-bd}, and credit constraints \citep{Simtowe2008-jn,Balana2020-hx} also play a role.\footnote{Complementary inputs include extension services and on-farm field trials, while complementary conditions refer to seed characteristics adapted for the terrain conditions and appropriate rainfall and temperature levels. Input availability and appropriate conditions have been found to be crucial to the adoption of improved maize in Tanzania \citep{Kaliba2000-jh}}. In the specific case of Ethiopia, labor, fertilizer use, farmers’ experience with extension packages, rainfall suitability and prices have been shown to increase the opportunity cost of not adopting improved wheat varieties \citep{Wale2006-bv}.\footnote{Market access and the accessibility of extension services are also critical in the adoption of improved chickpea varieties in Ethiopia \citep{Verkaart2019-ol}}

@@

These findings may also hold for our case, especially taking into consideration the degree of genetic dilution and dispersion of recycled or older-generation improved varieties, which are not as yield responsive to fertilizer application. In addition to this, prior evidence also indicates that farmers sub-optimally allocate fertilizer in response to their mislead beliefs about the crop variety (\citealt{euler2022because}). Such wrong beliefs or misclassification has also been shown to be driven by misperception instead of misreporting (\citealt{wossen2022misperceiving}), which is also the case for the farmers in our sample. These findings suggest that the seemingly small to negative returns to adoption that we find through the GRC model are driven by both a lower response to fertilizers of the older-generation improved varieties, but also to farmers’ sub-optimal application of fertilizers (and possibly other inputs) due to misperception.

The suitability of the improved varieties to weather conditions and the weather conditions themselves are two additional factors that may further explain the seemingly negative comparative advantages to adopting the misperceived improved varieties. Even moderate droughts have been shown to decrease maize yields of non-drought tolerant varieties by 20\% (\citealt{paul2021heterogeneous}). Additionally, although our results confirm the positive returns to DTM adoption (consistent with \citealt{wossen2017measuring}), even the DTM varieties distributed in Ethiopia are only resistant to mild drought but would also fail like any other non-drought tolerant variety in case of severe droughts.

>>>>

% 	\affil[1]{\small \emph{The World Bank}}
% 	\affil[2]{\small \emph{FAO}}
% 	\affil[3]{\small \emph{Assistant Research Professor at the Center for Data Science for Enterprise and Society, Cornell University}}
\author{%
 \textbf{Corresponding Author}: Aleksandr Michuda \footnote{Center for Data Science for Enterprise and Society, Cornell University, 237 Rhodes Hall, Ithaca, NY, USA, am2497@cornell.edu}%
 \and Cristina Chiarella\footnote{Earth and Life Institute, UCLouvain, 1348 Louvain-la-Neuve, Belgium}%
 \and Oscar Barriga-Cabanillas\footnote{The World Bank, 1818 H Street, NW Washington, DC USA}%
  \and Juan Sebastian Correa \footnote{Food and Agriculture Organization of the United Nations (FAO), 00153, Rome, Italy}%
  }

@@

    \clearpage\maketitle
	\thispagestyle{empty}
	\vspace*{-2em}
    \begin{center}\begin{abstract}
    \singlespacing
			\noindent  
			Despite large investments in developing new maize varieties, evidence shows that few farmers consistently adopt them. Moreover, average positive effects on yields mask considerable heterogeneity, particularly in contexts where seed are not traceable (missclasified as new varieties). We take a step further and study misclassification in the context of heterogeneous impacts to adoption. We use a three-year panel to estimate heterogeneous returns to adoption from self-reports of hybrid seed use and find null returns from adoption. We leverage unique information on maize DNA-fingerprinting to show this result is the result of bias in the estimation originating from  misclassification of seeds. Accounting for seed misclassification, we find that heterogeneity in the actual quality of the genetic material masks positive returns to adoption, with the highest returns found for farmers using higher-purity germplasm, drought-tolerant maize, and newly released varieties. Our findings point to the wide dispersion of older and genetically diluted varieties, for which poorer farmers may be paying a premium, and therefor the need for policies that expand the accessibility of improved, context and geography-specific seeds, which display higher yields.

@@

These include economic constraints, such as lack of liquidity and imperfections in credit markets \citep{Croppenstedt2003-pq; Carter2014-fm }, ambiguously defined property rights which limit investments \citep{Place2000-el} as well as supply constraints due to high transportation costs of agricultural inputs \citep{Byerlee2013-qk}, and seeds \citep{Bird2020-nt}. Other explanations include mismatches between seed characteristics and agricultural conditions \citep{Munshi2004-og} and lack of commitment devices \citep{Duflo2009-iv}.  
Alternative mechanism, closely related with our methodological approach is the misclassification of seed and seed genetic dilution (\citealt{wossen2022misperceiving}; \citealt{kosmowski2019varietal}). Yield improvement and disease resistance wanes over time as the genetic material is recycled \citep{Abate2015-rj}, making that the net returns from these \textit{improved} seed is lower than expected, and decreases in the long term as farmers continue to use genetically dilute seed varieties, or ones that have been recycled over many generations (\citealt{euler2022because}; \citealt{wossen2019estimating}). Moreover, as genes from improved varieties enter the regular seed market its inadvertent use can potentially lead to inferior yields since they require complementary inputs.

@@

We provide a way to account for misclassification of seed when additional, error-free data exists, such as DNA fingerprinting data. Our results contribute to the literature on the drivers of low adoption rates of improved seed varieties as we are the first to consider how heterogeneous impacts to seed adoption are affected by misclassification of seed. Considering that  missclassification introduces a non-classical measurement error that attenuates, or even reversed,  observed productivity impacts, the greater availability of DNA fingerprinting data makes possible a new revision of the role of misclassification of improved seed on the adoption of new varieties. For instance, \cite{wossen2019estimating} use DNA fingerprinting data to show that there is a 22\% bias in the productivity impact of improved cassava seed in Nigeria, suggesting that average impacts to adoption may not be as informative.  We estimate the comparative advantage of households directly using a Group Random Coefficients (GRC) Model and correct its heterogeneity estimates using misclassification rates from DNA fingerprinting data of the same plots \citep{Suri2011-oi, Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc}.

@@

The consensus in this literature is that the success of improved varieties depends on a range of factors, and that the widespread adoption of improved seeds is the result of complementary conditions and inputs that allow farmers to extract the full potential of these varieties, while other factos such as risk aversion \citep{Holden2016-vy}, farm size \citep{Ghimire2015-bd}, and credit constraints \citep{Simtowe2008-jn,Balana2020-hx} also play a role.\footnote{Complementary inputs include extension services and on-farm field trials, while complementary conditions refer to seed characteristics adapted for the terrain conditions and appropriate rainfall and temperature levels. Input availability and appropriate conditions have been found to be crucial to the adoption of improved maize in Tanzania \citep{Kaliba2000-jh}}. In the specific case of Ethiopia, labor, fertilizer use, farmers’ experience with extension packages, rainfall suitability and prices have been shown to increase the opportunity cost of not adopting improved wheat varieties \citep{Wale2006-bv}.\footnote{Market access and the accessibility of extension services are also critical in the adoption of improved chickpea varieties in Ethiopia \citep{Verkaart2019-ol}}

@@

These findings may also hold for our case, especially taking into consideration the degree of genetic dilution and dispersion of recycled or older-generation improved varieties, which are not as yield responsive to fertilizer application. In addition to this, prior evidence also indicates that farmers sub-optimally allocate fertilizer in response to their mislead beliefs about the crop variety leading to lower yield realization and consequent disadotpion  (\citealt{euler2022because}). Such wrong beliefs or misclassification has also been shown to be driven by misperception instead of misreporting (\citealt{wossen2022misperceiving}), which is also the case for the farmers in our sample. These findings suggest that the seemingly small to negative returns to adoption that we find through the GRC model are driven by a combination of lower responses to fertilizers of the older-generation improved varieties and farmers’ sub-optimal application of fertilizers (and possibly other inputs) due to misperception.

The suitability of the improved varieties to location-specific weather conditions as well as the variation in the weather conditions themselves are two additional factors that may further explain the seemingly negative comparative advantages to adopting the misperceived improved varieties. Even moderate droughts have been shown to decrease maize yields of non-drought tolerant varieties by 20\% (\citealt{paul2021heterogeneous}). Additionally, although our results confirm the positive returns to DTM adoption (consistent with \citealt{wossen2017measuring}), even the DTM varieties distributed in Ethiopia are only resistant to mild drought but would also fail like any other non-drought tolerant variety in case of severe droughts.
